---
title: "**Making your own OOP system in `R`**"
author: "Patrick Li"
institute: "Data Science Research Software Study Group (a.k.a. Brown Bag Session)"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css"]
    nature:
      highlightStyle: monokai
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
library(xaringanthemer)
style_solarized_dark(
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```


# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure

---

count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure

- `memory` to represent complex data structures

For example, this is how `R internal` accesses **integers** from an **integer vector**

<!-- This C macro is adopted from: r-source/src/include/Defn.h -->

```markdown
This C macro is adopted from: r-source/src/include/Defn.h

    Treat every 4 bytes of a memory block as a signed integer
                   ↓~~~~~↓
#define INTEGER(x) (int *) (((SEXPREC_ALIGN *) x) + 1)
                           ↑~~~~~~~~~~~~~~~~~~~~~~~~~↑
    Use pointer arithmetic to skip the header, e.g. vector length
    
                x        INTEGER(x)[0]   INTEGER(x)[1]
                ↓        ↓               ↓
               *--------*---------------*---------------*
Memomry Model: | Header | 4 bytes (int) | 4 bytes (int) | ...
               *--------*---------------*---------------*
```

---
count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure 

- `memory` to represent complex data structures
- `functions` to group code that performs a specific task together

A untested and inefficient Fibonacci number generator

```r
fib <- function(k) ifelse(k > 2, fib(k - 1) + fib(k - 2), k - 1) 
```

---

count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure 

- `memory` to represent complex data structures
- `functions` to group code that performs a specific task together
- `Loops` to perform operations repetitively over a container

Bubble sort

```r
x <- rnorm(100)
for (i in 1:(length(x) - 1)) 
  for (j in 1:(length(x) - i))
    if (x[j] > x[j + 1]) 
    {
      tmp_x <- x[j]
      x[j] <- x[j + 1]
      x[j + 1] <- tmp_x
    }
```

---

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

> There are no zero-cost abstraction. Abstractions have **run time**, **build time**, and **human costs**.

- Every `R` expression needs to be **parsed**, **compiled** and **interpreted** by the `R` Runtime. 

Every function call has a run time cost

```{r}
x <- y <- 2
foo <- function(x, y) x * y
microbenchmark::microbenchmark(x * y, foo(x, y), times = 10000)
```


```{r echo=FALSE}
rm(x, y)
```


---

count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

> There are no zero-cost abstraction. Abstractions have **run time**, **build time**, and **human costs**.

- Every `R` expression needs to be **parsed**, **compiled** and **interpreted** by the `R` Runtime. 
- Zero run time and build time costs doesn't necessary mean zero human costs.

---

count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

> There are no zero-cost abstraction. Abstractions have **run time**, **build time**, and **human costs**.

- Every `R` expression needs to be **parsed**, **compiled** and **interpreted** by the `R` Runtime. 
- Zero run time and build time costs doesn't necessary mean zero human costs.

> Each abstraction must provide more benefit than cost.

---

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure 

- `memory` to represent complex data structures
- `functions` to group code that performs a specific task together
- `Loops` to perform operations repetitively over a container

**OOP** to group instructions with the state they operate on?

---

count: false

# Abstraction

Inspired by the CppCon 2019 talk - Chandler Carruth “There Are No Zero-cost Abstractions”

In `R`, we use **abstraction** to describe data and data analysis procedure 

- `memory` to represent complex data structures
- `functions` to group code that performs a specific task together
- `Loops` to perform operations repetitively over a container

**OOP** to group instructions with the state they operate on?

Sure, as long as it produces more **human-readable code** with **acceptable overhead**.

---

# OOP

**OOP** stands for **object-oriented programming**, which is a programming paradigm based on **objects**. Usually, object is defined as a special type of data structure that can hold both **attributes** (data) and **methods** (associated behaviours). 

`R` has built-in **OOP** systems
- `S3`
- `S4`

There is another popular **OOP** system `R6` developed by Winston Chang. 

More details can be found in [Advanced R](https://adv-r.hadley.nz/oo.html). 

We will not focus on comparing them in this talk.

---

# `bandicoot`

```{r echo = FALSE}
library(bandicoot)
```

Some programming languages are fundamentally built upon **OOP** such as `Python`. 

--

`bandicoot` tries to provide a set of tools to build `Python-like` **OOP** system in `R`.

```r
devtools::install_github("TengMCing/bandicoot")
library(bandicoot)
```

--

- started in last year's Christmas holiday
- originally built for defining hierarchical structures of models used in visual inference experiments

--

### `bandicoot` used `environment` to emulate the **OOP** system

#### which is similar to `R6`, but built differerntly

---

# `environment`

Every `R` `function` associates with an `environment`. 

```{r}
environment(function(){})
```

--

An `environment` can reference itself with proper setup.

```{r}
env <- new.env()
env$self <- env #<<
identical(env, env$self)
```

--

A `function` can access its `enclosing environment` in the `function body`.

```{r}
env$foo <- function() self
environment(env$foo) <- env #<<
identical(env, env$foo())
```

---

# `environment`

To prevent function from directly accessing attributes other than `self`, an additional `environment` is needed with the `parent environment` to be the same.

```{r}
env <- new.env()
env$x <- 1
env$..method_env.. <- new.env(parent = parent.env(env))
env$..method_env..$self <- env

lobstr::tree(env)
```

--

```{r}
env$foo <- function() x
environment(env$foo) <- env$..method_env..

try(env$foo())
```

```{r echo = FALSE}
rm(foo, envir = env)
```


---

count: false

# `environment`

To prevent function from directly accessing attributes other than `self`, an additional `environment` is needed with the `parent environment` to be the same.

```r
env <- new.env()
env$x <- 1
env$..method_env.. <- new.env(parent = parent.env(env))
env$..method_env..$self <- env

lobstr::tree(env)
```

```{r echo = FALSE}
lobstr::tree(env)
```


```{r}
env$foo2 <- function() self$x
environment(env$foo2) <- env$..method_env..
try(env$foo2())
```

---

# `register_method()`

All this can be done by `register_method()`

```{r}
env <- new.env()
env$x <- 1
register_method(env, foo = function() self$x)

lobstr::tree(env)
```

---

# `register_method()`

The container and the self pointer name can be customized

```{r}
env <- new.env()
foo2 <- function() this$x + 1
register_method(env, foo_two = foo2, foo_three = foo2, 
                container_name = "container", self_name = "this")

lobstr::tree(env)
```

---

# `BASE` class

In `Python`, there is an `object` class, which provides essential attributes and methods for the OOP system (check [data model](https://docs.python.org/3/reference/datamodel.html) for more details).

In `bandicoot`, `BASE` class is the default `object` class, but you can write your own if you want advanced features.

```{r}
names(BASE)
```

---

# `new_class()`

`new_class()` is used for defining a new class, including new `object` class. 

```{r}
new_class(BASE, class_name = "CAT")
```


- `bandicoot`: **static method dispatch**. Method that needs to be called is decided at "build" time.
- `Python`: **dynamic method dispatch**. Method that needs to be called is decided at run time.

Primary concern is the overhead of **dynamic method lookup** and the difficulty of managing relationships between saved environments.

---

# Features of `BASE`

String representation of the object

```{r}
CAT <- new_class(BASE, class_name = "CAT")
CAT$..str..()
```

---

count: false

# Features of `BASE`

Parent class and the object class name

```{r}
CAT$..bases..
CAT$..type..
```

---

count: false

# Features of `BASE`

Class **constructor** and **initializer** (a **virtual function**)

```{r}
CAT$..new..()
CAT$..init..
```

---

count: false

# Features of `BASE`

Class instantiation 

```{r}
little_cat <- CAT$instantiate()
little_cat$..str..()
```

"Official" String representation of the object

```{r}
little_cat$..repr..()
```

---

# Workflow

This is a simple `STAFF` class defined in `Python`

```python
class STAFF(object):
    
    def __init__(self, name, age):
        self.name = name
        self.age = age
    
    def get_email(self):
        return f"ʕ•́ᴥ•̀ʔっ♡{self.name}@company.com"
```

---

# Workflow

Define a class with class description (a class factory)

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) {
  
  new_class(BASE, env = env, class_name = "STAFF")
  
  init_ <- function(name, age) {
    self$name <- name
    self$age <- age
  }
  
  get_email_ <- function() {
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com")
  }
  
  register_method(env, ..init.. = init_, get_email = get_email_)
  
  return(env)
}

STAFF <- class_STAFF()
```


---

# Workflow

#### 1. New an environment `env` if it is not provided

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { #<<
  
  new_class(BASE, env = env, class_name = "STAFF")
  
  init_ <- function(name, age) {
    self$name <- name
    self$age <- age
  }
  
  get_email_ <- function() {
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com")
  }
  
  register_method(env, ..init.. = init_, get_email = get_email_)
  
  return(env)
}

STAFF <- class_STAFF()
```

---

# Workflow

#### 2. New a class `STAFF` in the environment `env`

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { 
  
  new_class(BASE, env = env, class_name = "STAFF") #<<
  
  init_ <- function(name, age) {
    self$name <- name
    self$age <- age
  }
  
  get_email_ <- function() {
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com")
  }
  
  register_method(env, ..init.. = init_, get_email = get_email_)
  
  return(env)
}

STAFF <- class_STAFF()
```

---

# Workflow

#### 3. Define the initializer `..init..` to capture `name` and `age`

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { 
  
  new_class(BASE, env = env, class_name = "STAFF") 
  
  init_ <- function(name, age) { #<<
    self$name <- name            #<<
    self$age <- age              #<<
  }                              #<<
  
  get_email_ <- function() {
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com")
  }
  
  register_method(env, ..init.. = init_, get_email = get_email_)
  
  return(env)
}

STAFF <- class_STAFF()
```


---

# Workflow

#### 4. Define a method `get_email` to get the email address

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { 
  
  new_class(BASE, env = env, class_name = "STAFF") 
  
  init_ <- function(name, age) { 
    self$name <- name            
    self$age <- age              
  }                              
  
  get_email_ <- function() {              #<<
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com") #<<
  }                                       #<<
  
  register_method(env, ..init.. = init_, get_email = get_email_)
  
  return(env)
}

STAFF <- class_STAFF()
```

---

# Workflow

#### 5. Register methods

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { 
  
  new_class(BASE, env = env, class_name = "STAFF") 
  
  init_ <- function(name, age) { 
    self$name <- name            
    self$age <- age              
  }                              
  
  get_email_ <- function() {              
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com") 
  }                                       
  
  register_method(env, ..init.. = init_, get_email = get_email_) #<<
  
  return(env)
}

STAFF <- class_STAFF()
```

---

# Workflow

#### 6. Return the class/environment

```{r}
class_STAFF <- function(env = new.env(parent = parent.frame())) { 
  
  new_class(BASE, env = env, class_name = "STAFF") 
  
  init_ <- function(name, age) { 
    self$name <- name            
    self$age <- age              
  }                              
  
  get_email_ <- function() {              
    glue::glue("ʕ•́ᴥ•̀ʔっ♡{self$name}@company.com") 
  }                                       
  
  register_method(env, ..init.. = init_, get_email = get_email_) 
  
  return(env) #<<
}

STAFF <- class_STAFF()
```

---

# Workflow

#### 7. Build an instance of class `STAFF`

```{r}
Patrick <- STAFF$instantiate(name = "Patrick", age = 18)
```

#### 8. Get name, age, and email of the object

```{r}
Patrick$name
Patrick$age
Patrick$get_email()
```

---

# More realistic examples

$$y_i = x_i + e_i, \quad e_i \sim N(0, 1+x_i^2), \quad i=1,...,n.$$

```{r}
library(visage)
z <- rand_normal(mu = 0, sigma = 1)
x <- rand_uniform(-1, 1)
e <- closed_form(~sqrt(1 + x^2) * z)
y <- closed_form(~x + e)
y
```


---

# More realistic examples

```{r}
y$gen(5, rhs_val = TRUE) |> y$as_dataframe()
```


<!-- --- -->

<!-- #### Please check [some of the classes I wrote](https://github.com/TengMCing/visage/tree/master/R) and the [reference page](https://tengmcing.github.io/visage/reference/index.html). -->

<!-- #### So far I have written 10 classes for my visual inference research using this system. It works well for me and it is very easy to debug since the design is simple. -->
